---
title: "amber-file"
author: "Amber Potter"
date: '2022-11-14'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## For Reference:

Mini translation of variable names:
- rdaysfromindex = days from the date of the first catheterization for that
patient that shows significant coronary artery disease (CAD) that is recorded in the Duke Health System
- RSEQCATHNUM = this variable records the sequential number for the current catheterization
record
- rsubjecID = identifier, take out
- YRCATH_G = year the procedure took place?
- demographic variables = need levels to be relabeled
- ACS = Acute coronary syndrome, coronary artery disease symptom status at the time of presentation
- CHFSEV = Congestive heart failure is defined as evidence of fluid retention due to cardiac causes. There should be a history of one or more of the following: exertional dyspnea, orthopnea, paroxysmal nocturnal dyspnea, rales, pulmonary congestion on x-ray, or a ventricular gallop. This variable measures SEVERITY of CHF
- DPCABG = Days from this catheterization to the closest preceding CABG.
- DPMI = Days from this catheterization to the closest preceding myocardial infarction. 
- DPPCI = Days from this catheterization to the closest preceding Percutaneous Coronary Intervention.
- HXANGINA = History of anginal pain ever (excluding myocardial infarction pain)
- HXCEREB = History of a stroke, a transient ischemic attack, hemiplegia, carotid surgery or stenting. An asymptomatic bruit is not included in the definition
- HXCHF = History of congestive heart failure not due to acute MI
- HXCOPD= History of chronic obstructive pulmonary disease (COPD)
- HXDIAB = history of diabetes not differentiating between type I and type II
- HXHTN = hypertension history
- HXHYL = Previous diagnosis and/or treatment of hypercholesterolemia by a physician.
- HXMI = History of a prior myocardial infarction
- HXSMOKE = history of smoking
- NUMPRMI = number of MRIs in history
- DIASBP_R = diastolic blood pressure
- PULSE_R = heart rate
- SYSBP_R = systolic blood pressure
- CBRUITS = presence of carotid bruits 
- HEIGHT_R = height
- S3 = third heart sound detected?
- WEIGHT_R = weight 
- CREATININE_R = Serum Creatinine (mg/dL) (as reported)
- HDL_R = High Density Lipid (mg/dL) (as reported)
- LDL_R = Low Density Lipid (mg/dL) (as reported)
- TOTCHOL_R = Total Cholesterol (mg/dL) (as reported)




### Packages

```{r}
library(tidyverse)
library(survival)
```

## Data

```{r}
cathr <- readRDS("dukecath_pre-processed.RDS")

# Overview of patient information
cathr$ACS <- factor(cathr$ACS, levels=c(0,1,2,3,4), labels=c("No ACS", "STEMI", "NSTEMI", "MI Unspecified", "UA"))
cathr$GENDER <- factor(cathr$GENDER, levels=c(0,1), labels=c("M", "F"))
cathr$RACE_G <- factor(cathr$RACE_G, levels=c(1,2,3), labels=c("Caucasian", "African American", "Other"))
cathr$INTVCATH <- factor(cathr$INTVCATH, levels=c(0,1), labels=c("No", "Yes"))


cathr_clean <- cathr %>% 
  filter(ACS %in% c("NSTEMI", "UA")) %>%  
  filter(DIAGCATH == 0)
```


```{r}



# 7080 rows left
cathr_clean <- cathr %>% 
  filter(ACS %in% c("Non-STEMI", "Unstable Angina")) %>%  
  filter(DIAGCATH == 0) # 18.4%
```

```{r}
# treatment
cathr_clean %>% 
  count(INTVCATH)
```
```{r}
# Percentage of column that is NA.
(colMeans(is.na(cathr_clean)))*100
```

```{r}
summary(cathr_clean$DPMI)
```

- Days to Closest Previous Coronary Artery Bypass Surgery (DPCABG)
- Days to Closest Previous Myocardial Infarction (DPMI)
- Days to Closest Previous Percutaneous Coronary Intervention (DPPCI)

```{r}
ggplot(data = cathr_clean, aes(x = DPMI)) + 
  geom_histogram(binwidth = 2) + 
  labs(x = "DPMI", 
       y = "Count", 
       title = "Distribution of Days to Closest Previous Myocardial Infarction")
```
# what is an interesting treatment variable?

Treatment: Whether PCI is performed (INTVCATH)
Response: (DAYS2LKA)

```{r}
library(survival)
# install.packages("survminer")
library(survminer)

coxm1 <- coxph(Surv(DAYS2LKA, DEATH) ~ INTVCATH + 
                 , data = cathr_clean)

summary(coxm1)
```

# Filter data for NSTEMI and UA

this is the data we are working on

```{r}
cathr_modified <- cathr_modified %>%
  filter(ACS == "NSTEMI" |ACS == "UA") %>%
  filter(TYPE_SUBSEQUENT != "Both (same day)")
```



## InitialAFT Model


- time between diagnostic and interventional cath in days (two different models)
- indicator for if diagnostic and interventional cath on same day (two different models)
- race variables
- NOT gender

```{r}
cathr_modified_aft <- cathr_modified %>% 
  filter(DSINTV > 0) %>%
  filter(!is.na(DSINTV), !is.na(DEATH))


library(survival)
initial_aft_e <-  survreg(Surv(DSINTV, DEATH) ~ DSINTV_CAT +
               DSINTV +
               TYPE_SUBSEQUENT + 
                 GENDER +
                 RACE_G +
                 HXDIAB, 
                 data = cathr_modified_aft, dist = "exponential")
summary(initial_aft_e)
```

```{r}
initial_aft_w <- survreg(Surv(DAYS2LKA, DEATH) ~ DSINTV_CAT +
               DSINTV +
               TYPE_SUBSEQUENT + 
                 GENDER +
                 RACE_G +
                 HXDIAB, 
                 data = cathr_modified_aft, dist = "weibull")
summary(initial_aft_w)
```

```{r}
initial_aft_n <- survreg(Surv(DAYS2LKA, DEATH) ~ DSINTV_CAT +
               DSINTV +
               TYPE_SUBSEQUENT + 
                 GENDER +
                 RACE_G +
                 HXDIAB, 
                 data = cathr_modified_aft, dist = "lognormal")
summary(initial_aft_n)
```

```{r}
initial_aft_l <- survreg(Surv(DAYS2LKA, DEATH) ~ DSINTV_CAT +
               DSINTV +
               TYPE_SUBSEQUENT + 
                 GENDER +
                 RACE_G +
                 HXDIAB, 
                 data = cathr_modified_aft, dist = "loglogistic")
summary(initial_aft_l)
```

# Cox model: how does same day relate to not same day?

```{r}
coxm_bin <- coxph(Surv(DAYS2LKA, DEATH) ~ DSINTV_CAT +
               TYPE_SUBSEQUENT + 
                 GENDER +
                 RACE_G +
                 HXDIAB, data = cathr_modified)

summary(coxm_bin)
```

```{r}
library(tidyverse)

cathr <- read.csv("dukecathr.csv")

# Overview of patient information
cathr$ACS <- factor(cathr$ACS, levels=c(0,1,2,3,4), labels=c("No ACS", "STEMI", "NSTEMI", "MI Unspecified", "UA"))
cathr$GENDER <- factor(cathr$GENDER, levels=c(0,1), labels=c("M", "F"))
cathr$RACE_G <- factor(cathr$RACE_G, levels=c(1,2,3), labels=c("Caucasian", "African American", "Other"))
cathr$INTVCATH <- factor(cathr$INTVCATH, levels=c(0,1), labels=c("No", "Yes"))
```


```{r}
# treatment
cathr_clean %>% 
  count(INTVCATH)
```
```{r}
# Percentage of column that is NA.
(colMeans(is.na(cathr_clean)))*100
```

```{r}
summary(cathr_clean$DPMI)
```

Days to Closest Previous Coronary Artery Bypass Surgery (DPCABG)
Days to Closest Previous Myocardial Infarction (DPMI)
Days to Closest Previous Percutaneous Coronary Intervention (DPPCI)

```{r}
ggplot(data = cathr_clean, aes(x = DPMI)) + 
  geom_histogram(binwidth = 2) + 
  labs(x = "DPMI", 
       y = "Count", 
       title = "Distribution of Days to Closest Previous Myocardial Infarction")
```

```{r}
data <- readRDS('dukecath_pre-processed.RDS')
data_clean <- data %>% 
  filter(ACS %in% c("NSTEMI", "UA")) %>% 
  filter(TYPE_SUBSEQUENT != "Both (same day)")
```


Variables of Interest: DSINTV and DSINTV_CAT.
Response: Death and DAYS2LKA.

```{r}
hist(data_clean$DSINTV[data_clean$DSINTV > 0 & data_clean$DSINTV < 50])
```

```{r}
library(survival)
# install.packages("survminer")
library(survminer)
```

```{r}
colnames(data_clean)
```

```{r}
coxm1 <- coxph(Surv(DAYS2LKA, DEATH) ~
               log(DSINTV + 0.1) +
               TYPE_SUBSEQUENT + 
                 # demographic
                 GENDER +
                 RACE_G +
                 # patient history and co-morbidities
                 HXMI +
                 HXCEREB +
                 HXDIAB + 
                 HXSMOKE + 
                 # lab results
                 TOTCHOL_R +
                 # Catheterization results
                 NUMDZV +
                 # interaction terms
                 NUMDZV*TYPE_SUBSEQUENT +
                 HXMI*TYPE_SUBSEQUENT +
                 HXCEREB*TYPE_SUBSEQUENT +
                 HXDIAB*TYPE_SUBSEQUENT
                 , data = data_clean)

summary(coxm1)
```




# Binary multiple days

```{r}
coxm1 <- coxph(Surv(DAYS2LKA, DEATH) ~
               TYPE_SUBSEQUENT + 
                 # demographic
                 GENDER +
                 RACE_G +
                 # patient history and co-morbidities
                 HXMI +
                 HXCEREB +
                 HXDIAB + 
                 HXSMOKE + 
                 # lab results
                 #TOTCHOL_R +
                 # Catheterization results
                 NUMDZV +
                 # interaction terms
                 NUMDZV*TYPE_SUBSEQUENT +
                 HXMI*TYPE_SUBSEQUENT +
                 HXCEREB*TYPE_SUBSEQUENT +
                 HXDIAB*TYPE_SUBSEQUENT
                 , data = data_clean)

summary(coxm1)
```


```{r echo = FALSE}
library(broom)
library(knitr)
tidy_coxm1 <- tidy(coxm1)

exp_model_coefs <- as_tibble(exp(tidy_coxm1$estimate))

conf_int <- confint(coxm1)
model_conf_int_2.5 <- as_tibble(exp(conf_int))[,1]
model_conf_int_97.5 <- as_tibble(exp(conf_int))[,2]

model_variables <- as_tibble(c(names(coxm1$coefficients)))


model_pvals <- as.data.frame(coef(summary(coxm1))[,5])[,1]

Variables <- c("Intercept", "First Overtime", "Second Overtime", "Third Overtime", "Fourth Overtime", "Star Committing", "Superstar Committing", "Star Disadvantaged", "Superstar Disadvantaged", "Attendance", "Home Team Committing", "Seconds Left", "2016-17", "2017-18", "2018-19", "2019-20", "2020-21", "2021-22", "Playoff Game", "Offensive Foul", "Personal Foul", "Shooting Foul", "Ball Handling Foul", "Star Committing:Star Disadvantaged", "Superstar Committing:Star Disadvantaged", "Star Committing:Superstar Disadvantaged", "Superstar Committing:Superstar Disadvantaged")

model_output_full <- data.frame(model_variables, 
           exp_model_coefs, 
           model_conf_int_2.5,
           model_conf_int_97.5, 
           model_pvals) %>%
  kable(digits = 3, col.names = c("Variables", "Odds Ratio Coefficient", "2.5% CI for Odds Ratio", "97.5% CI for Odds Ratio", "P-Value (alpha = 0.05"))





SE <- sqrt( diag( coxm1$se ))

exp_coef <- exp(coxm1$coef)

exp_coef <- coxm1$confint

coef[2]


## conf intervals for fixed effects
se <- sqrt(diag(vcov(coxm1)))
# table of estimates with 95% CI
tab <- cbind(Est = fixef(coxm1), se, LL = fixef(coxm1) - 1.96 * se, UL = fixef(coxm1) + 1.96 *
    se)

df <- data.frame(tab)

df <- df %>%
  mutate(exp_Est = exp(Est),
         exp_LL = exp(LL),
         exp_UL = exp(UL),
         z = Est/se,
         p = ifelse(z>0, 2*pnorm(-z), 2*pnorm(z))) %>%
  dplyr::select(Est, se, exp_Est, exp_LL, exp_UL, p) %>%
  mutate(p = case_when(p<0.0005 ~ "<0.001",
                       TRUE ~ as.character(formatC(round(p, 3), format='f', digits=3))))

df %>%
  kable(digits=3, col.names = c("Estimate", "Std. Error", "exp(Estimate)", "exp(Lower Bound", "exp(Upper Bound)", "P-value (alpha = 0.05"))


```

```{r include = FALSE, eval=FALSE}
library(jpeg)
jpeg(file="ggcoxdiag1.jpeg")
coxdiag1 <- ggcoxdiagnostics(coxm1, type = "schoenfeld", title = "Diagnostic Plot for Type of Subsequent Procedure Model")
dev.off()


jpeg(file="ggcoxdiag2.jpeg")
coxdiag2 <- ggcoxdiagnostics(coxm2, type = "schoenfeld", title = "Diagnostic Plot for Days to Intervention Model")
dev.off()

```

```{r include = FALSE, eval=FALSE}
img1 <- readJPEG(source = "ggcoxdiag1.jpeg",  native = FALSE)
img2 <- readJPEG(source = "ggcoxdiag2.jpeg",  native = FALSE)
plot(0:1,0:1,type="n",ann=FALSE,axes=FALSE)
rasterImage(img1,0,0,1,1)
plot.new() 
rasterImage(img2,0,0,1,1)

# ggcoxdiagnostics(coxm1, type = "deviance", linear.predictions=F)
```